Lessons learned from bugs in models of human history

### 2020-05-20 paper.tex
Minor language tweaks

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ required users to specify cryptic command
line options to describe demographic models.
Particularly for the large models we use today, these are not
intuitive or comprehensible for humans.
The Python API for \msprime\, by contrast, allows the user to
state models in a more human-readable manner: 
Example of model specification for the two softwares is provided in the Methods section. 
%\sgcomment{Other software, such as SLiM, 
%have developed advanced user interfaces and even dedicated programming languages 
%to facilitate the specification of various modes of evolution.\cite{}
%}

@@

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. However, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime~ver.~0.3.0 and remained in the documentation for
around four years.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
After these initial checks were made, however, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations 
relative to the intended model.

@@

First, the design of user interface and API for scientific software matters,
and many bugs can be prevented by using more intuitive interfaces.
Whereas the original \msprime\ API required the user to pass
three separate parameters to specify a demographic model, \msprime~ver.~1.0
will introduce a Demography class, which wraps these three parameters.
Thus, instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
with the error-inducing need to re-enter parameter information,
we would now write
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

@@

Graphical inspection of demographic models can also help identify issues, especially if 
visualization can be automated. msprime does contain a demography debugging tool summarizing 
demographic events occurring over time, and various efforts are underway to facilitate 
the specification and visualization of population genetics models \cite{zhou2018popdemog}, \sgcomment{Mention Aaron's code and Slim?}.

>>>>

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is the Python API that is its primary interface, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ required users to specify cryptic command
line options to describe demographic models.
Particularly for the large models we use today, these are not
intuitive or comprehensible for humans.
The Python API for \msprime\, by contrast, allows the user to
state models in a more human-readable manner: 
an examplle of model specification for the two softwares is provided in the Methods section. 
%\sgcomment{Other software, such as SLiM, 
%have developed advanced user interfaces and even dedicated programming languages 
%to facilitate the specification of various modes of evolution.\cite{}
%}

@@

To illustrate the demography API, \msprime\ included a description of a widely-used
three population Out-of-Africa model~\citep{gutenkunst2009inferring}
as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past,
followed by a more recent split of European and Asian populations, with variable rates of
continuous migration between each of the populations. Regrettably, the implementation in the
\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
populations, when there should have been just a single randomly mating population, migration was
allowed to occur between the ancestral population and a second population with size equal to
the Eurasian bottleneck size for all time into the past
(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
the tutorial for \msprime~ver.~0.3.0 and remained in the documentation for
around four years.

@@

In another publication using this model~\citep{martin2017human},
a separate error was introduced: the model itself was defined as suggested
in the documentation (using updated parameters
from~\citet{gravel2011demographic}) and inspected
using the \msprime\ debugging tools.
Despite these initial checks being made, the simulation
was performed without passing the parameter of demographic events,
so that the three populations
never merged as expected and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
leading to a vast overestimate of the divergence across human populations.
While the correct model predicts a mean $F_{ST}$ of
$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
ranging between $0.3 - 0.6$, depending on the populations considered.
Overall diversity was also strongly affected: heterozygosity was more than
doubled in African populations and reduced by more than half in Eurasian populations 
relative to the intended model.

@@

First, the design of user interface and API for scientific software matters,
and many bugs can be prevented by using more intuitive interfaces.
Whereas the original \msprime\ API required the user to pass
three separate parameters to specify a demographic model, \msprime~ver.~1.0
will introduce a Demography class, which wraps these three parameters.
Instead of writing
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
\end{lstlisting}
with the error-inducing need to re-enter parameter information,
we would now write
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  migration_matrix=migration_matrix,
  demographic_events=demographic_events)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}

@@

Graphical inspection of demographic models can also help identify issues, especially if 
visualization can be automated. \msprime\ does contain a demography debugging tool summarizing 
demographic events occurring over time, and various efforts are underway to facilitate 
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.
